---
title: "Smoking, Age and Death"
geometry: margin=1in
output:
  html_document:
    df_print: paged
header-includes:
- \usepackage{fontspec}
- \setmainfont{Times New Roman}
fontsize: 12pt
---

Smoking is a major health concern among the population, however many individuals of numerous age groups continue to smoke. The goal is to analyze potential relationships between age group, smoking status and mortality rate among women.

 



Let's load and view the data:

```{r smoking-1, message=FALSE, warning=FALSE}

library(tidyverse)

#Load data from the library
library(SMPracticals)
data(smoking)

smoking_tbl <- as_data_frame(smoking)
dplyr::glimpse(smoking_tbl)

```

Now we need a way to view correlations between variables before fitting any models, and in this case a table is most suitable.

```{r table-1, message=FALSE, warning=FALSE}

# Relationship between mortality and smoking
xtabs(cbind(dead,alive) ~ smoker,data=smoking_tbl) %>% prop.table(1)

```

Looking at potential relationships, the first table shows that a greater proportion of smokers in the study were alive after 20 years than non-smokers.

Let's try a binomial regression model:

```{r model-1, message=FALSE, warning=FALSE}

# Binomial regression, mortality against smoking
smoking_glm1 <- glm(cbind(dead,alive) ~ smoker, data = smoking_tbl, family = binomial)
summary(smoking_glm1)

```

The *binomial regression model* for mortality against smoking shows a significant negative relationship between the variables, which indicates that smoking decreases mortality rate. This is unexpected, but another factor (age) has not been taken into account, which could explain this unusual relationship. Also, the residual deviance is quite large compared to its degrees of freedom, so this model is not a good fit.

To investigate this unintuitive relationship, we display a different table to show the relationship between smoking and age in groups of `dead` or `alive`. 

```{r table-2, message=FALSE, warning=FALSE}

# Relationship between smoking and age in two groups: dead or alive.
xtabs(cbind(dead,alive) ~ smoker + age,data=smoking_tbl) %>% prop.table(2)

```

In this table, there is a larger proportion of younger women who smoke, relative to older women who smoke. Many of these younger women who smoke were still alive after 20 years into the study, while many of the older women passed away.


Now try another *binomial regression model*, but with age groups as a predictor:

```{r model-2, message=FALSE, warning=FALSE}

# Binomial regression, mortality against age and smoking
smoking_glm2 <- glm(cbind(dead,alive) ~ age + smoker, data = smoking_tbl, family = binomial)
summary(smoking_glm2)

```

This model is a very strong fit since the residual deviance is quite small relative to its degrees of freedom. Now that `age` has been accounted for, the `smoker` variable is positively correlated with mortality; this is an example of Simpson's paradox. The dependence of smoking status and mortality rate are explained by their respective relationship with age (i.e. smoking and mortality are dependent, conditional on age).

If investigators in this study did not measure age, they may have incorrectly concluded that smoking correlates with a lower risk of death. In observational studies such as this one, investigators need to be careful in drawing conclusions before considering other factors that can influence relationships between the variables of interest.
