---
title: "STA442 Homework 1"
author: "Kevin Dang - 1003205079"
date: "Due 24 September 2019"
output: pdf_document
---

# Fruit Flies

Fruit flies have relatively short lifespans, which makes them a good candidate for experiments that involve collecting data on their lifetimes and other factors that may affect their lifetimes, including sexual activity and thorax length. Lifespan is measured in days, and we are not told what units thorax length is measured in however it is likely to be very small units such as millimeters. Sexual activity is divided into five categories, with 25 male fruit flies in each group: fly kept alone (`isolated`), one pregnant female (`activityone`), eighty pregnant females (`activitymany`), one virgin female (`activitylow`), and eight virgin females (`activityhigh`). The first three groups are the control groups which involve no mating while the latter two categories with virgin female fruit flies involve mating. 

The model that is used to answer our questions regarding fruit fly lifetimes is a Gamma generalized linear model. In Gamma regression, the data follow a Gamma distribution with range parameter $\phi = \mu_i/\nu$ and shape parameter $\nu$, i.e. $Y_i \sim Gamma(\mu_i/\nu)$ with $E(Y_i)=\mu_i$. We are using a log-link so we have a model of: 
\[\log(\mu_i)=X_i\beta=\beta_{intercept}+\beta_{I(activity)}X_1+\beta_{thorax}X_2\] 
Table 1 below contains the output for the Gamma regression model, and Table 2 contains the 95\% confidence intervals for the parameters. The intercept represents flies that were kept solitary, and this will be our baseline. Starting with $\beta_1$ (`activityone`), we have a $100(\exp(0.055)-1)=5.7\%$ chance of increase in lifespan if the fruit fly is kept with one pregnant female. If the fruit fly is kept with one virgin female (`activitylow`), then it represents a lifespan decrease of -11.0\% ($100(\exp(-0.116)-1)=-11.0\%$). Next is `activitymany`, also known as flies kept with eight pregnant females, which has an increase of 8.5\% (see previous calculations). Finally, we have a decrease in longevity by 34.0\% for `activityhigh`. The p-values for `activityone` and `activitymany` are quite large compared to the standard 0.05 significance level so these results are not significant. This makes sense because those two groups are also control groups which involve no mating. Thorax length appears to have the greatest effect, because it increases lifespan by a whopping 1370\%. In Figure 1 we have a histogram with a Gamma density line as a visual aid to help us check the model fit. The red Gamma line follows the histogram shape quite well so the Gamma generalized linear model is a good fit to the fruit fly data.


```{r fly, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
data('fruitfly', package='faraway')

fruitfly$length <- fruitfly$thorax - median(fruitfly$thorax)
flyglm <- glm(longevity ~ activity+length, family=Gamma(link='log'), data=fruitfly)
knitr::kable(summary(flyglm)$coef, digits=3, caption="Gamma GLM, lifetimes as a function of the thorax length and activity") # Table 1
```

\newpage

```{r fly2, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(confint(flyglm), digits=3, caption="95% Confidence Intervals for Fly Gamma GLM ") # Table 2
```


```{r fly3, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Histogram of Fruit Fly lifespan with density of a Gamma distribution',fig.height=4,fig.width=5,fig.show='hold'}
shape = 1/summary(flyglm)$dispersion
hist(fruitfly$longevity, prob=TRUE,
     xlab='days', main = "Lifetimes of Fruit Flies")
xSeq = seq(par('usr')[1], par('usr')[2], len=200)
lines(xSeq, 
      dgamma(xSeq, shape=shape, 
             scale = exp(flyglm$coef['(Intercept)'])/shape),
      col='red', lwd=2
)
```


## Fruit Flies Summary
A scientific laboratory conducted a study on how the level of sexual activity and thorax length affects the lifetimes of male fruit flies. Thorax length was measured for each fly and it had quite a large effect, increasing the probability of a longer than average lifespan by 1370\%. Sexual activity also had an impact on the lifespans. The male fruit flies who were kept with one pregnant female or eight pregnant females are more likely to slightly live longer than the average isolated fruit fly, by 5.7\% and 8.5\% respectively, however these values may be due to random chance as the model showed that there was no significant difference. These groups are not able to mate so they are less sexually active. However the groups which were kept with one virgin female or eight virgin females have on average a decrease in lifetime by 11\% and 34\% respectively. These two groups are more sexually active, particularly the second group. In conclusion, fruit flies with lower sexual activity levels tend to live longer compared to the groups who have very high sexual activity levels.

\newpage
# Appendix A

## Flies Code
```{r flycode, echo=TRUE, message=FALSE, warning=FALSE, results='hide', fig.show='hide'}
library(tidyverse)
data('fruitfly', package='faraway')

fruitfly$length <- fruitfly$thorax - median(fruitfly$thorax)
flyglm <- glm(longevity ~ activity+length, family=Gamma(link='log'), data=fruitfly)
knitr::kable(summary(flyglm)$coef, digits=3, 
             caption="Gamma GLM, lifetimes as a function of the thorax length and activity") # Table 1

knitr::kable(confint(flyglm), digits=3, 
             caption="95% Confidence Intervals for Fly Gamma GLM ") # Table 2

# Code below borrowed from cars.r
shape = 1/summary(flyglm)$dispersion
hist(fruitfly$longevity, prob=TRUE,
     xlab='days', main = "Lifetimes of Fruit flies")
xSeq = seq(par('usr')[1], par('usr')[2], len=200)
lines(xSeq, 
      dgamma(xSeq, shape=shape, 
             scale = exp(flyglm$coef['(Intercept)'])/shape),
      col='red', lwd=2
)
```

\newpage

# Smoking

## Introduction
The 2014 American National Youth Tobacco Survey is a rich data source that will allow us to answer various research hypotheses. The first problem that we want to investigate is whether the regular use of chewing tobacco, snuff or dip among three specific ethnic groups in America is different. The ethnic groups are European-American, Hispanic-American, and African-American. In this analysis, we must account for whether they live in an urban or rural area. The second hypothesis to investigate involves determining how likely it is for two youths of the opposite sex with similar characteristics to have used a hookah or waterpipe at least once. A sample of the data that will be used in the analysis is in Table 3 below.

```{r smoke, echo=FALSE, message=FALSE, warning=FALSE}
dataDir = "../data"
smokeFile = file.path(dataDir, "smokeDownload.RData")
if (!file.exists(smokeFile)) {
download.file("http://pbrown.ca/teaching/appliedstats/data/smoke.RData",
smokeFile)
}
load(smokeFile)
knitr::kable(smoke[1:5,c('Age','Sex','RuralUrban','Race','chewing_tobacco_snuff_or', 'ever_tobacco_hookah_or_wa')], caption="Smoking Data") # Table 3
smokeSub = smoke[smoke$Age >= 10, ] # 9 yr olds are suspicious
```

## Methods
The method used to tackle the first hypothesis is a logistic regression model. Looking at Tables 4 and 5 below, we can see that the potential predictors for our model are binary, which makes sense to use logistic regression in this scenario. The model is as follows: $Y_i\sim Bin(N_i,\mu_i)$ with $\log(\frac{\mu_i}{1-\mu_i})=X_i\beta$. $Y_i$ is the number of people who have used chewing tobacco, snuff or dip on 1 or more days in the past 30 days and $N_i$ is the number of observations. The probability of a person chewing tobacco, snuff or dip given covariates $X_i$ (race, urban/rural location) is $\mu_i$. The parameter of interest is $\beta_{Race}$ and the confounders are $\beta_{RuralUrban}$ and $\beta_{intercept}$. Notice that in Table 4 the proportion of European-Americans who have tried chewing tobacco is largest, but in Table 5 the percentage of European-Americans who live in rural areas is also the highest. Chewing tobacco is more common in rural areas so we must account for that in the model. The model will be fit via a two way interaction: 
\[\log(\frac{\mu_i}{1-\mu_i})= \beta_{intercept}+\beta_{Race}X_1+\beta_{RuralUrban}X_2+\beta_{Race}\beta_{RuralUrban}X_3\]

For the second research hypothesis we also fit a logistic regression model. This time, $Y_i$ represents the number of people who ever smoked tobacco out of a hookah or waterpipe. So the probability of a person who has smoked tobacco out of a hookah or waterpipe given $X_i$ (sex, age, race) is $\mu_i$. The parameter of interest is $\beta_{Sex}$, while the confounders are $\beta_{Age},\beta_{Race},\beta_{RuralUrban},\beta_{intercept}$. The model is as follows:
\[\log(\frac{\mu_i}{1-\mu_i})= \beta_{intercept}+\beta_{Sex}X_1+\beta_{Age}X_2+\beta_{Race}X_3+\beta_{RuralUrban}X_4\]

```{r table4, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(table(smokeSub$Race, smokeSub$chewing_tobacco_snuff_or), caption="Chewing tobacco, snuff or dip, by Race") # Table 4
```

\newpage
```{r table5, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(table(smokeSub$Race, smokeSub$RuralUrban), caption="Urban vs Rural, by Race") # Table 5
```

## Results
The logistic regression model for chewing tobacco is shown in Table 6 below along with the 95\% confidence intervals in Table 7. The intercept represents European-Americans who live in an urban areas which is our reference. The odds ratio of an African-American in an urban area using chewing tobacco, snuff or dip regularly is exp(-1.006)=0.366, which means that the their odds are 100(exp(-1.006)-1)=-63.4\% lower compared to the odds for urban European-Americans. For urban Hispanic-Americans the odds decrease is -30.7\% (see previous calculation). If a European-American lives in a rural area then the percentage odds of using chewing tobacco, snuff or dip regularly increases by 220.9\%. For a rural African-American the percentage odds changes by -50.6\% and for rural Hispanic-Americans the percentage decrease is -48.1\%. The most noticeable difference here is the huge increase in odds by more than three times for European-Americans in rural areas compared to urban areas. This is due to the fact chewing tobacco is a rural phenomenon. There was also an increase for the African-Americans, however the case is different for Hispanic-Americans who saw a decrease in likelihood. Perhaps Hispanic-Americans who live in rural areas happen to be more watchful over their children compared to those in the busy cities because of the fact that tobacco is more common there, but we need more data to confirm this. From these numbers, we can see that every group with the exception of rural European-Americans is less likely to have used chewing tobacco relative to the urban European-Americans. While accounting for urban and rural areas, we can see that regular use of chewing tobacco is more common amongst European-Americans than for Hispanic-Americans and African-Americans.


```{r glm1, echo=FALSE, message=FALSE, warning=FALSE}
tobaccoglm <- glm(chewing_tobacco_snuff_or ~ Race*RuralUrban,data=smokeSub,family=binomial(link="logit"))
knitr::kable(summary(tobaccoglm)$coef, digits=3, 
             caption="Logistic Regression Model, Regular use of Chewing Tobacco") # Table 6
```

\newpage
```{r table7, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(confint(tobaccoglm), digits=3, 
             caption="95% Confidence Intervals for Tobacco GLM ") # Table 7
```

The logistic regression model for hookah or waterpipe is shown in Table 8 below along with the 95\% confidence intervals in Table 9. The parameter of interest corresponds is `SexF`, which says that holding all other demographic characteristics fixed, there is a 100(exp(0.042)-1)=4.3% increase in the odds of using a hookah or waterpipe if the person is female. The p-value of 0.327 is quite large and the increase in odds is rather small so there is no significant difference between two very similar individuals of the opposite sex in the likelihood of using a hookah or waterpipe.

```{r glm2, echo=FALSE, message=FALSE, warning=FALSE}
hookahglm <- glm(ever_tobacco_hookah_or_wa ~ Sex+Age+Race+RuralUrban,data=smokeSub,family=binomial(link="logit"))
knitr::kable(summary(hookahglm)$coef, digits=3, 
             caption="Logistic Regression Model, Use of hookah or waterpipe") # Table 8
knitr::kable(confint(hookahglm), digits=3, 
             caption="95% Confidence Intervals for Hookah GLM") # Table 9
```

## Smoke Summary
From this study, we conclude that regular use of chewing tobacco is more common amongst European-Americans than for Hispanic-Americans and African-Americans. An interesting takeaway from the results is that European-Americans in rural areas are more than three times as likely to use chewing tobacco than European-Americans in urban areas. The study also found that given two individuals of the opposite sex with very similar demographic characteristics such as age, race and location, the likelihood of having used a hookah or waterpipe at least once is the same.

# Appendix B

## Smoking Code
```{r smokecode, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}
# Load in Data
dataDir = "../data"
smokeFile = file.path(dataDir, "smokeDownload.RData")
if (!file.exists(smokeFile)) {
download.file("http://pbrown.ca/teaching/appliedstats/data/smoke.RData",
smokeFile)
}
load(smokeFile)

knitr::kable(smoke[1:5,c('Age','Sex','Grade','RuralUrban','Race','chewing_tobacco_snuff_or',
                         'ever_tobacco_hookah_or_wa')], caption="Smoking Data") # Table 3
smokeSub = smoke[smoke$Age >= 10, ] # 9 yr olds are suspicious, remove

knitr::kable(table(smokeSub$Race, smokeSub$chewing_tobacco_snuff_or), 
             caption="Chewing tobacco, snuff or dip, by Race") # Table 4
knitr::kable(table(smokeSub$Race, smokeSub$RuralUrban), 
             caption="Urban vs Rural, by Race") # Table 5

# Binomial GLM Regular use of Chewing Tobacco.
smokeglm1 <- glm(chewing_tobacco_snuff_or ~ Race*RuralUrban,data=smokeSub,
                 family=binomial(link="logit"))
knitr::kable(summary(tobaccoglm)$coef, digits=3, 
             caption="Logistic Regression Model, Regular use of Chewing Tobacco") # Table 6
knitr::kable(confint(tobaccoglm), digits=3, 
             caption="95% Confidence Intervals for Tobacco GLM ") # Table 7

# Binomial GLM Likelihood of hookah or waterpipe
smokeglm2 <- glm(ever_tobacco_hookah_or_wa ~ Sex+Age+Race+RuralUrban,data=smokeSub,
                 family=binomial(link="logit"))
knitr::kable(summary(hookahglm)$coef, digits=3, 
             caption="Logistic Regression Model, Likelihood of hookah or waterpipe") # Table 8
knitr::kable(confint(hookahglm), digits=3, 
             caption="95% Confidence Intervals for Hookah GLM") # Table 9
```
